Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-05-07 setup.tex

 \subsubsection{Rotation and vertical shear profiles}\label{vshear_def}
The equilibrium rotation field, $\Omega(r,z)$ follows from the density field and centrifugal balance (Eq.\ \ref{hori_equilibrium}),
giving 
\begin{align}\label{eqm_rot}
  \frac{\Omega^2(r,z)}{\OmK^2(r)}=1 +
  \frac{p+q}{\Gamma} h^2(r) 
  +\frac{s}{\Gamma} \left(1-\frac{1}{\sqrt{1+z^2/r^2}}\right), 
\end{align}
for all $\Gamma$, where $s\equiv q+p(1-\Gamma)$.  We also refer to the
epicyclic frequency $\kappa$ defined via
\begin{align}\label{kap2_def}
\kappa^2 = \frac{1}{r^3}\frac{\p}{\p r}\left(r^4\Omega^2\right). 
\end{align}

@@

\subsection{Entropy gradients and vertical buoyancy}
The gradients of entropy, $S\equiv
C_p\ln{P^{1/\gamma}/\rho}$, in our disk models are
\begin{subequations}
\begin{align}
  &\frac{\p S}{\p r} = C_p\left[\frac{s}{\gamma r} +
    \left(\frac{\Gamma}{\gamma}-1\right)\frac{\p \ln{\rho}}{\p
       r}\right],\\
  &\frac{\p S}{\p z} = C_p
  \left(\frac{\Gamma}{\gamma}-1\right)\frac{\p \ln{\rho}}{\p z} = {C_p g_z \over c_s^2} {\Gamma - \gamma \over \gamma},\label{dSdz}
\end{align} 
\end{subequations}
 where $C_p$ is the heat capacity at
constant pressure.

The vertical buoyancy frequency $N_z$ is 
\begin{align}
  N_z^2 \equiv - C_p^{-1}g_z \frac{\p S}{\p z} = {g_z^2 \over c_s^2}{\gamma - \Gamma \over \gamma}. %c_s^2 \left(\frac{\gamma -
   %   \Gamma}{\gamma}\right)\left(\frac{\p\ln\rho}{\p z}\right)^2,  
    \label{nzsq_def}
\end{align}
We only consider convectively stable disks with with $\Gamma <
\gamma$ and  $N_z^2\geq0$. % , and the disk is 
% convectively stable in the vertical direction.

\subsection{Stability without cooling}\label{solberg}
In the absence of cooling, axisymmetric stability is ensured
if both Solberg-Hoiland criteria  are satisfied:
\begin{align}
  \kappa^2 - \frac{1}{\rho C_p}\nabla P \cdot \nabla S &> 0,\\
  -\frac{\p P}{\p z}\left(\frac{\p j^2}{\p r}\frac{\p S}{ \p z} -
    \frac{\p j^2}{\p z}\frac{\p S}{\p r} \right)&>0. \label{second_SH} 
\end{align}
\citep{tassoul78}, where $j=r^2\Omega$ is the specific angular
momentum and $\kappa = r^{-3} \p_r j^2 $ is the epicyclic frequency.  For rotationally-supported, thin disks %  pressure gradients are small 
% compared to rotation, so 
the first criterion is generally satisfied. 
Thus, to assess stability, we consider the second
criterion. 
Without loss of generality, we evaluate Eq. \ref{second_SH} in $z>0$
where $\p_zP<0$. Here we approximate $\p_r j^2\simeq r^3\OmK^2$ and
use Eq. \ref{vertical_shear} for $\p_zj^2$. Then 
\begin{align}\label{solberg2}
  % \gamma - \Gamma > \frac{s h^2}{\Gamma}
  % \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s
  %   -\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]. 
  \gamma - \Gamma > \frac{ h^2}{\Gamma}
  \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s^2
    -s\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]
\end{align} 
implies stability. For typical model parameters the right hand
side (RHS) is $O( h^2)$. Hence, instability in the adiabatic limit  
requires $\gamma < \Gamma + O( h^2)$. Since we consider
$ h\ll1$, this implies the equilibrium disk must have a small
entropy gradient in order to be unstable to axisymmetric adiabatic
perturbations. This is generally not true for passively irradiated, stably
stratified disks \citep{chiang97}, in which the VSI requires rapid
cooling, as shown below.

>>>>

 \subsubsection{Rotation and vertical shear profiles}\label{vshear_def}
The equilibrium rotation field, $\Omega(r,z)$ follows from the density field and centrifugal balance (Eq.\ \ref{hori_equilibrium}),
giving 
\begin{align}\label{eqm_rot}
  \frac{\Omega^2(r,z)}{\OmK^2(r)}=1 +
  \frac{p+q}{\Gamma} h^2(r) 
  +\frac{s}{\Gamma} \left(1-\frac{1}{\sqrt{1+z^2/r^2}}\right), 
\end{align}
for all $\Gamma$, where $s\equiv q+p(1-\Gamma)$.  We also refer to the
epicyclic frequency $\kappa$ defined via
\begin{align}\label{kap2_def}
\kappa^2 \equiv \frac{1}{r^3}\frac{\p}{\p r}\left(r^4\Omega^2\right) = \frac{1}{r^3}\frac{\p j^2}{\p r}, 
\end{align}
where $j$ is the specific angular momentum.

@@

\subsection{Entropy gradients and vertical buoyancy}
The gradients of entropy, $S\equiv
C_P\ln{P^{1/\gamma}/\rho}$, in our disk models are
\begin{subequations}
\begin{align}
  &\frac{\p S}{\p r} = C_P\left[\frac{s}{\gamma r} +
    \left(\frac{\Gamma}{\gamma}-1\right)\frac{\p \ln{\rho}}{\p
       r}\right], \label{dSdr}\\
  &\frac{\p S}{\p z} = C_P
  \left(\frac{\Gamma}{\gamma}-1\right)\frac{\p \ln{\rho}}{\p z} = {C_P g_z \over c_s^2} {\Gamma - \gamma \over \gamma},\label{dSdz}
\end{align} 
\end{subequations}
 where $C_P$ is the heat capacity at
constant pressure.

The vertical buoyancy frequency $N_z$ is 
\begin{align}
  N_z^2 \equiv - C_P^{-1}g_z \frac{\p S}{\p z} = {g_z^2 \over c_s^2}{\gamma - \Gamma \over \gamma}. %c_s^2 \left(\frac{\gamma -
   %   \Gamma}{\gamma}\right)\left(\frac{\p\ln\rho}{\p z}\right)^2,  
    \label{nzsq_def}
\end{align}
We only consider convectively stable disks with with $\Gamma <
\gamma$ and  $N_z^2\geq0$. % , and the disk is 
% convectively stable in the vertical direction.

\subsection{Stability without cooling}\label{solberg}
In the absence of cooling, axisymmetric stability is ensured
if both Solberg-Hoiland criteria  are satisfied:
\begin{align}
  \kappa^2 - \frac{1}{\rho C_P}\nabla P \cdot \nabla S &> 0,\\
  -\frac{\p P}{\p z}\left(\frac{\p j^2}{\p r}\frac{\p S}{ \p z} -
    \frac{\p j^2}{\p z}\frac{\p S}{\p r} \right)&>0. \label{second_SH} 
\end{align}
\citep{tassoul78}. %, where $j=r^2\Omega$ is the specific angular
%momentum. %and $\kappa = r^{-3} \p_r j^2 $ is the epicyclic frequency.  
For rotationally-supported, thin disks %  pressure gradients are small 
% compared to rotation, so 
the first criterion is generally satisfied. 
Thus, to assess stability, we consider the second
criterion. 
Without loss of generality, we evaluate Eq. \ref{second_SH} in $z>0$
where $\p_zP<0$. Here we approximate $\p_r j^2\simeq r^3\OmK^2$ and
use Eq. \ref{vertical_shear} for $\p_zj^2$. Then 
\begin{align}\label{solberg2}
  % \gamma - \Gamma > \frac{s h^2}{\Gamma}
  % \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s
  %   -\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]. 
  \gamma - \Gamma > \frac{ h^2}{\Gamma}
  \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s^2
    -s\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]
\end{align} 
implies stability. For typical model parameters the right hand
side (RHS) is $O( h^2)$. Hence, instability in the adiabatic limit  
requires $\gamma < \Gamma + O( h^2)$. Since we consider
$ h\ll1$, this implies the equilibrium disk must have a small
entropy gradient in order to be unstable to axisymmetric adiabatic
perturbations. This is generally not true for passively irradiated, stably
stratified disks \citep{chiang97}, in which the VSI requires rapid
cooling, as shown below.
